\Title

### 2013-02-03 synmut.tex
fig.1 refers to p1

\end{abstract}
\maketitle

\section{Introduction}

@@

\section{Results}

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p8.pdf}
\label{fig:aftsyn}}\\
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p8.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of allele frequencies in \env, C2-V5, from
patient 8~\cite{shankarappa_consistent_1999},
divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}

One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. It has
been suggested early on that parts of the viral genome that has the potential to
form stems is better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

@@

\subsection{Deleterious mutations are brought to high frequency by hitchhiking}

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.

@@

\section{Discussion}
Despite several known functional roles for RNA secondary structure in the HIV
genome, synonymous mutations are often used as approximately neutral markers in
evolutionary studies of viruses. By analyzing the fate of mutations in
longitudinal data of HIV \env{} evolution, we have shown that the majority of
synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
Comparison with recent biochemical studies of binding propensity of bases in RNA
genome suggest that these mutations are deleterious, at least in part, because they disrupt
stems in RNA secondary structures. Furthermore, we provide evidence that these
mutations are brought to high frequency through linkage to adaptive mutations.

@@

\section{Methods}
\subsection{Sequence data collection}

Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
showed signs of HIV compartimentalization into subpopulations and were
discarded; a total of 11 patients with approximately 6 time points each and 10
sequences per time point were eventually analyzed. The time interval or
resolution between two consecutive sequences was approximately 6 to 18 months.

\comment{How did you determine who to discard. Maybe we should include the PCA
plots into the supplement.}

\subsection{Sequence analysis}
The good sequences were aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
consensus nucleotide sequence at the first time point was used to classify
alleles as ancestral or derived at all sites. Problematic sites that included large frequencies of gaps
were excluded from the analysis to avoid artefactual substitutions due to
alignment errors. Time series of allele frequencies were extracted from the
sequences. \comment{The alignments used are available for download}.

The synonymity of a mutation was assigned if the rest of the codon was
in the ancestral state and using the standard genetic code. Cases where more
than one mutation within the codon was observed were discarded. Slightly
different criteria for synonymous/nonsynonymous discrimination yielded similar
results.

\subsection{Fixation probability and secondary structure}
For the estimate of times to fixation/extinction, polymorphisms were
binned by frequency and the time to reaching the first boundary (fixation or
extinction) was stored. For the fixation probability, the long-time limit of the
resulting curves was used, excluding polymorphisms that arose late in the
clinical history (and would have had no time to reach either boundary).

For the correlation analysis with RNA secondary structure, the SHAPE scores were
downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
of the alignment of the longitudinal sequences with the reference used by
\citet{watts_architecture_2009}, SHAPE reactivities were assigned to most sites.
Problematic assignments in indel-rich regions were excluded from the analysis.
In order to restrict the analysis to synonymous polymorphisms, a lower frequency
threshold of 0.15 was used (other thresholds yielded the same results). Since
very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
contains much more points than the ``fixed" one.

The V loops and flanking regions were identified manually starting from the
annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
similar approach was used to label the C2-V5 region sequenced in
ref.~\citep{shankarappa_consistent_1999}.

\subsection{Computer simulations}
Simulations were performed using the recently published software
FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations were performed and yielded comparable results. For each
set of parameters, approximately 100 simulation runs were averaged over. In each
run, a random fitness landscape with specified statistical properties (e.g.
density of beneficial sites, average deleterious effect of synonymous changes) was generated.
Although the curves shown in \FIG{simfixpvar} are not very smooth, small
parameter changes resulted in overall consistent trends across many repetitions.

For the discussion of simulation parameters, the areas below or above the neutral
diagonal were estimated from the binned fixation probabilities using the linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes, because the HIV data are quite scarse themselves.

\comment{All analysis scripts are available for download\ldots.}. 
\section*{Acknowledgements}
We are grateful for stimulating discussions with Jan Albert and Trevor Bedford.
This work is supported by the ERC starting grant HIVEVO 260686.

>>>>

\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env~, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of allele frequencies in \env, C2-V5, from
patient 1~\cite{shankarappa_consistent_1999},
divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. It has
been suggested early on that parts of the viral genome that has the potential to
form stems is better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Despite several known functional roles for RNA secondary structure in the HIV
genome, synonymous mutations are often used as approximately neutral markers in
evolutionary studies of viruses. By analyzing the fate of mutations in
longitudinal data of HIV \env{} evolution, we have shown that the majority of
synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
Comparison with recent biochemical studies of binding propensity of bases in RNA
genome suggest that these mutations are deleterious, at least in part, because they disrupt
stems in RNA secondary structures. Furthermore, we provide evidence that these
mutations are brought to high frequency through linkage to adaptive mutations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences are collected for published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
show signs of HIV compartimentalization into subpopulations and are
discarded (see supplementary materials);
a total of 11 patients with approximately 6 time points each and 10
sequences per time point are eventually analyzed. The time interval or
resolution between two consecutive sequences is approximately 6 to 18 months.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences are aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
consensus nucleotide sequence at the first time point is used to classify
alleles as ancestral or derived at all sites. Problematic sites that include large frequencies of gaps
are excluded from the analysis to avoid artefactual substitutions due to
alignment errors. Time series of allele frequencies are extracted from the
sequences. \comment{The alignments used are available for download}.

The synonymity of a mutation is assigned if the rest of the codon was
in the ancestral state and using the standard genetic code. Cases where more
than one mutation within the codon is observed are discarded. Slightly
different criteria for synonymous/nonsynonymous discrimination yield similar
results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimate of times to fixation/extinction, polymorphisms are
binned by frequency and the time to reaching the first boundary (fixation or
extinction) is stored. For the fixation probability, the long-time limit of the
resulting curves is used, excluding polymorphisms that arise late in the
clinical history (hence would have no time to reach either boundary).

For the correlation analysis with RNA secondary structure, the SHAPE scores are
downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
of the alignment of the longitudinal sequences with the reference used by
\citet{watts_architecture_2009}, SHAPE reactivities are assigned to most sites.
Problematic assignments in indel-rich regions are excluded from the analysis.
In order to restrict the analysis to synonymous polymorphisms, a lower frequency
threshold of 0.15 was used (other thresholds yield the same results). Since
very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
contains much more points than the ``fixed" one.

The V loops and flanking regions are identified manually starting from the
annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
similar approach is used to label the C2-V5 region sequenced in
ref.~\citep{shankarappa_consistent_1999}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Simulations are performed using the recently published software
FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations are performed and yield comparable results. For each
set of parameters, approximately 100 simulation runs are averaged over. In each
run, a random fitness landscape with specified statistical properties (e.g.
density of beneficial sites, average deleterious effect of synonymous changes) is generated.

For the discussion of simulation parameters, the areas below or above the neutral
diagonal were estimated from the binned fixation probabilities using linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Methods availability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All analysis scripts are available for download.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We are grateful for stimulating discussions with Jan Albert and Trevor Bedford.
This work is supported by the ERC starting grant HIVEVO 260686.
